---
title: "Decoding Biases in AI - Group Project"
author: "mp"
date: '2022-11-29'
output: html_document
---

## Results and Discussion

We begin our initial investigation by conducting exploratory data analysis, where we visualize how *all* videos combined that followed from a left-leaning starting point compare to those associated with a right-leaning starting point, regardless of the position in the recommendation chain.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Loading libraries
library(tidyverse)
library(ggplot2)
library(readxl)
# Loading dataset
yt_data <- read_excel("~/Desktop/yt_data.xlsx")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Re-coding data
yt_data_candidate <- yt_data %>%
  rename(Index = candidate_support)
yt_data_candidate$Index <- as.character(yt_data_candidate$Index)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Visualizing data
video_plot_candidate <- ggplot(data = yt_data_candidate, aes(x = starting_video, fill = Index)) + xlab("Political leaning of starting video") + ylab("Video count") + labs(title = "Figure 1: Video support for election-denying candidates/entities, by starting point") + geom_bar(position = "stack") + scale_fill_manual(values = c("cadetblue2", "khaki2", "coral1")) + theme_classic()
video_plot_candidate
```

As Figure 1. above shows, the number of videos *criticizing* election deniers (coded as -1) produced by the left-leaning starting point is greater than the number of videos *supporting* election deniers (coded as 1) that followed from the right-leaning starting point. Further, the proportion of neutral videos (coded as 0) appears to be greater for the right-leaning starting video than for the left-leaning starting video, suggesting that more conservative content is *less* likely to steer the users towards videos supporting election deniers than the more liberal content is towards videos supporting election *non*-deniers; this pattern is also similar for videos that directly support or criticize election fraud conspiracy theories, as presented in Figure 2. below.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Re-coding data
yt_data_claim <- yt_data %>%
  rename(Index = claim_support)
yt_data_claim$Index <- as.character(yt_data_claim$Index)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Visualizing data
video_plot_claim <- ggplot(data = yt_data_claim, aes(x = starting_video, fill = Index)) + xlab("Political leaning of starting video") + ylab("Video count") + labs(title = "Figure 2: Video support for election-denying claims, by starting point") + geom_bar(position = "stack") + scale_fill_manual(values = c("cadetblue2", "khaki2", "coral1")) + theme_classic()
video_plot_claim
```

As shown above, the number of videos supporting election-denying claims that followed from the right-leaning starting point is substantially smaller than the proportion of videos criticizing election-denying claims that followed from the left-leaning starting point, further indicating that left-leaning content, rather than right-leaning content, might be more likely to produce biased videos.

This initial analysis, however, does not take into account differences between recommendation chains. For this reason, it might be helpful to evaluate how the results compare between different "cold starts."

The table below presents the mean measure of support for election-denying candidates/entities (which we will call the "candidate index") among different recommendation chains (cold starts) that followed from the left-leaning starting video.


```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Creating new datasets
yt_data_L_candidate <- filter(yt_data, starting_video == "L")
yt_data_R_candidate <- filter(yt_data, starting_video == "R")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Wrangling left/candidate dataset
yt_data_L_candidate <- yt_data_L_candidate %>%
  group_by(cold_start) %>%
  summarize(
    candidate_index = mean(candidate_support)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Displaying left/candidate dataset
yt_data_L_candidate
```

As the table above shows, most recommendation chains generated by the left-leaning starting point and composed of 6 videos, not surprisingly, have the index below 0, indicating that the recommendation algorithm overwhelmingly steered the user clicking through a recommendation chain towards content criticizing election-denying candidates or entities. However, there are some anomalies, such as recommendation chain #1, #4, and #8. The pattern is similar for the mean measure of support for election-denying claims, called "claim index," as presented in the table below.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Creating new datasets
yt_data_L_claim <- filter(yt_data, starting_video == "L")
yt_data_R_claim <- filter(yt_data, starting_video == "R")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Wrangling left/claim dataset
yt_data_L_claim <- yt_data_L_claim %>%
  group_by(cold_start) %>%
  summarize(
    claim_index = mean(claim_support)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Displaying left/claim dataset
yt_data_L_claim
```

The same indexes for recommendation chains that started from the right-leaning starting point, however, show a slightly different pattern, as presented in the two tables below.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Wrangling right/candidate dataset
yt_data_R_candidate <- yt_data_R_candidate %>%
  group_by(cold_start) %>%
  summarize(
    candidate_index = mean(candidate_support)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Displaying right/candidate dataset
yt_data_R_candidate
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Wrangling right/claim dataset
yt_data_R_claim <- yt_data_R_claim %>%
  group_by(cold_start) %>%
  summarize(
    claim_index = mean(claim_support)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Displaying right/claim dataset
yt_data_R_claim
```

Here, all recommendation chains composed of 6 videos that followed from the right-leaning starting point have the mean support for election-denying candidates/entities above 0, indicating that, overall, these chains, unlike those for left-leaning starting point, might have produced more biased content. For election-denying claims, however, the right-leaning video-produced chains appear to be more neutral, with many chains having the claim index of 0. 

However, when we average all recommendation chain indexes for left-leaning and right-leaning starting points, separately, the results show that chains originating from the left-leaning starting video have roughly the same index as chains originating from the right-leaning starting video in absolute terms. The claim index average remains slightly higher for the left-leaning starting point, again, in absolute terms.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Producing total means
yt_data_candidate_index <- yt_data %>%
  group_by(cold_start) %>%
  group_by(starting_video) %>%
  summarize(
    candidate_index = mean(candidate_support)
  )
yt_data_claim_index <- yt_data %>%
  group_by(cold_start) %>%
  group_by(starting_video) %>%
  summarize(
    claim_index = mean(claim_support)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
yt_data_candidate_index
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
yt_data_claim_index
```

As such, the overall results appear to be mixed and thus did not produce enough evidence to support the claim that the YouTube recommendation algorithm steers its users towards far-right content, defined as content supporting candidates or claims in favor of election fraud conspiracy theories,  more aggressively than it does towards content criticizing said theories. In fact, the results appear to be more suggestive of the opposite.
