---
title: "Decoding Biases in AI - Group Project"
author: "mp"
date: '2022-11-29'
output: html_document
---

## Results and Discussion

We begin our initial investigation by conducting exploratory data analysis, where we visualize how *all* videos combined that followed from a left-leaning starting point compare to those associated with a right-leaning starting point, regardless of the position in the recommendation chain.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Loading libraries
library(tidyverse)
library(ggplot2)
library(readxl)
# Loading dataset
yt_data <- read_excel("~/Desktop/yt_data.xlsx")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Re-coding data
yt_data_candidate <- yt_data %>%
  rename(Index = candidate_support)
yt_data_candidate$Index <- as.character(yt_data_candidate$Index)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Visualizing data
video_plot_candidate <- ggplot(data = yt_data_candidate, aes(x = starting_video, fill = Index)) + xlab("Political leaning of starting video") + ylab("Count") + labs(title = "Figure 1: Video support for election-denying candidates/entities, by starting point") + geom_bar(position = "stack") + scale_fill_manual(values = c("cadetblue2", "khaki2", "coral1")) + theme_classic()
video_plot_candidate
```

As Figure 1. above shows, the number of videos *criticizing* election deniers (coded as -1) produced by the left-leaning starting point is greater than the number of videos *supporting* election deniers (coded as 1) that followed from the right-leaning starting point. Further, the proportion of neutral videos (coded as 0) appears to be greater for the right-leaning starting video than the left-leaning starting video, suggesting that more conservative content is *less* likely to steer the users towards videos supporting election deniers than the more liberal content is towards videos supporting election *non*-deniers; this pattern is also similar for videos that directly support or criticize election fraud conspiracy theories, as presented in Figure 2. below.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Re-coding data
yt_data_claim <- yt_data %>%
  rename(Index = claim_support)
yt_data_claim$Index <- as.character(yt_data_claim$Index)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Visualizing data
video_plot_claim <- ggplot(data = yt_data_claim, aes(x = starting_video, fill = Index)) + xlab("Political leaning of starting video") + ylab("Count") + labs(title = "Figure 2: Video support for election-denying claims, by starting point") + geom_bar(position = "stack") + scale_fill_manual(values = c("cadetblue2", "khaki2", "coral1")) + theme_classic()
video_plot_claim
```

As shown above, the number of videos supporting election-denying claims that followed from the right-leaning starting point is substantially smaller than the proportion of videos criticizing election-denying claims that followed from the left-leaning starting point, further indicating that left-leaning content, rather than right-leaning content, might be more likely to produce biased videos.

This initial analysis, however, does not take into account differences between recommendation chains. For this reason, it might be helpful to evaluate how the results compare between different "cold starts."






```{r}
yt_data_L <- filter(yt_data, starting_video == "L")
yt_data_R <- filter(yt_data, starting_video == "R")
```


```{r}
yt_data_L <- yt_data_L %>%
  group_by(cold_start) %>%
  summarize(
    political_index = mean(candidate_support)
  )
```

```{r}
yt_data_R <- yt_data_R %>%
  group_by(cold_start) %>%
  summarize(
    political_index = mean(candidate_support)
  )
```


```{r}
yt_data_index <- yt_data %>%
  group_by(cold_start) %>%
  group_by(starting_video) %>%
  summarize(
    political_index = mean(candidate_support)
  )
```




